A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-17 bioc-workflow-resubmit.Rnw
changes until sml, with travis condif update

If the protein naming between the marker sets and the \texttt{MSnSet}
dataset are different e.g. the markers are labelled by Uniprot
accession numbers and the dataset entries are labelled by Uniprot
entry names, one will have to convert and match the proteins according
to the appropriate identifier. Sometimes, we find the equivalent entry
name, Uniprot ID or accession number is stored in the feature
metadata, which makes conversion between identifers relatively
straightforward. If this is not the case however, conversion can be
performed using \Biocpkg{biomaRt}, the Bioconductor
\href{http://bioconductor.org/help/workflows/annotation/Annotation_Resources/}{annotation
  resouces} or any conversion softwares available online.

@@

Finally, in addition to \texttt{plot2D}, the \texttt{plot3D} function
allows to interactively explore a 3-dimensional plot of the data.

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object and then use the \textit{highlightOnPlot} function to
visualise these. For example, if we wanted to highlight
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and
Q8C156, which are some of the proteins known to form part of the
13S condensin complex, we would call the following code:

@@

Users can also create several sets of \texttt{FeaturesOfInterest}
object and store them in a \texttt{FoICollection}.

@@

With the aim of maximising the sub-cellular resolution and,
consequently, the reliability in protein sub-cellular assignments, we
follow the advice in \cite{Trotter:2010} and combine replicated spatial
proteomics experiments as described above. Indeed, Trotter et
al. have shown a significant improvement in protein–organelle
association upon direct combination of single experiments, in
particular when these resolve different subcellular niches.

@@

The more relevent comparison unit is not a single channel, but rather
the complete protein occupancy profiles, which are best visualised
experiment-wide on a PCA plot. As such, we prefer to focus on the
direct, qualitative comparison of individual replicate PCA plots,
assuring that each displays acceptable sub-cellular resolution. Note
that in the code chunk below, we mirror the x-axis to represent the
two figures with the same orientation. The interactive "compare"
application part of the \Biocpkg{pRolocGUI} package is also useful for
examining replicate experiments (see the next section interactive
visualisation for details).

@@

Visualisation and data exploration is an important aspect of data
analyses allowing one to shed light on data structure and patterns of
interest. Using the \Biocpkg{pRolocGUI} package we can interactively
visualise, explore and interrogate quantitative spatial proteomics
data. The \Biocpkg{pRolocGUI} package is currently under active
development and it relies on the \texttt{shiny} framework for
reactivity and interactivity. The package currently distributes 3
different GUI's (\textit{main} (default), \textit{compare} or
\textit{classify}) which are wrapped and launched by the
\texttt{pRolocVis} function.

@@

\subsubsection*{The classify application}
The final application \textit{classify}, has been designed to view
machine learning classification results according to user-specified
thresholds for the assignment of proteins to its sub-cellular
location, as discussed later in the subsection \textit{thresholding}
in the \textit{supervised machine learning section}.

@@

In addition to extracting new phenotypes, novelty detection methods
are also useful for confirming the presence of known or postulated
clusters in an unbiased fashion. For example, in \cite{hyper} the
\texttt{phenoDisco} algorithm was used to confirm the data-specific presence
of the nucleus and nucleus sub-compartments. In the code chunk below,
we demonstrate how to do this analysis, highlighting some of the
optional arguments and parameters available for phenotype extraction
and give some advice on how to interpret the output.

As the \texttt{phenoDisco} algorithm is semi-supervised it uses both labelled
(markers) and unlabelled data to explore the data structure and find
new sub-cellular data clusters. Thus the first step is to define some
input labelled data i.e. markers, that the algorithm will use as input
for the supervised learning aspect of the algorithm. As described in
\cite{hyper} we define a set of markers to use as input for the analyses
that cover well-known residents from three distinct organelle
structures; the mitochondria, plasma membrane and ER, and from three
well-known and abundant protein complexes; the proteasome and two
ribosomal subunits, 40S and 60S.  These input markers are stored in
the \texttt{featureData} column of \texttt{hl} where \texttt{fcol = "phenoDisco.Input"}. We
can use the convenience accessor function \texttt{getMarkers} to print out a
table of the markers contained in this marker set. These initial
markers were manually curated using information from the UniProt
database, the Gene Ontology and the literature.

@@

The argument \texttt{times} indicates the number of times we run
unsupervied Gaussian Mixture Modelling before defining a new phenotype
cluster.  The recommended minimum and default value is 100. In the
above code chunk we increase the value to \texttt{times = 200} as we
have found for larger datasets (e.g. 5000+ proteins) a higher
\texttt{times} is requried for convergence. \texttt{GS} defines the
minimum number of proteins allowed per new data cluster and thus
heavily influences what type of new clusters are extracted. For
example, if a user is interested in the detection of small complexes
they may wish to use a small \texttt{GS = 10}, or \texttt{= 20} etc.
If they wish to detect larger, more abundant sub-cellular niches a
much higher \texttt{GS} would be preferable. Specifying a small
\texttt{GS} can be more time consuming than using a larger
\texttt{GS}, and there is a trade off between finding interesting
small complexes and those that may not be of interest as we find there
is a tendancy to find more noise when using a small \texttt{GS}
compared to using a higher one.

>>>>

If the protein naming between the marker sets and the \texttt{MSnSet}
dataset are different e.g. the markers are labelled by Uniprot
accession numbers and the dataset entries are labelled by Uniprot
entry names, one will have to convert and match the proteins according
to the appropriate identifier. Sometimes, we find the equivalent entry
name, Uniprot ID or accession number is stored in the feature
metadata, which makes conversion between identifers relatively
straightforward. If this is not the case however, conversion can be
performed using \Biocpkg{biomaRt}, the Bioconductor
\href{http://bioconductor.org/help/workflows/annotation/Annotation_Resources/}{annotation
  resources} or any conversion softwares available online.

@@

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object and then use the \textit{highlightOnPlot} function to
visualise these. For example, if we wanted to highlight
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and
Q8C156, which are some of the proteins known to form part of the
13S condensin complex, we would call the code displayed
on Figure \ref{fig:foi}. Users can also create several sets of \texttt{FeaturesOfInterest}
object and store them in a \texttt{FoICollection}.

@@

@@

With the aim of maximising the sub-cellular resolution and,
consequently, the reliability in protein sub-cellular assignments, we
follow the advice in \cite{Trotter:2010} and combine replicated spatial
proteomics experiments as described above. Indeed, Trotter \textit{et
  al.} have shown a significant improvement in protein–organelle
association upon direct combination of single experiments, in
particular when these resolve different subcellular niches.

@@

The more relevent comparison unit is not a single channel, but rather
the complete protein occupancy profiles, which are best visualised
experiment-wide on a PCA plot. As such, we prefer to focus on the
direct, qualitative comparison of individual replicate PCA plots,
assuring that each displays acceptable sub-cellular resolution. Note
that in the code chunk below, we mirror the x-axis to represent the
two figures with the same orientation. The interactive "compare"
application part of the \Biocpkg{pRolocGUI} package is also useful for
examining replicate experiments (see the next section
\textit{Interactive visualisation} for details).

@@

Visualisation and data exploration is an important aspect of data
analyses allowing one to shed light on data structure and patterns of
interest. Using the \Biocpkg{pRolocGUI} package, we can interactively
visualise, explore and interrogate quantitative spatial proteomics
data. The \Biocpkg{pRolocGUI} package relies on the \texttt{shiny}
framework for reactivity and interactivity. It distributes 3 different
GUI's (\textit{main} (default), \textit{compare} or \textit{classify})
which are wrapped and launched by the \texttt{pRolocVis} function.

@@

\subsubsection*{The classify application}
The final application \textit{classify}, has been designed to view
machine learning classification results according to user-specified
thresholds for the assignment of proteins to its sub-cellular
location, as discussed later in the subsection \textit{Thresholding}
in the \textit{Supervised machine learning} section.

@@

In addition to extracting new phenotypes, novelty detection methods
are also useful for confirming the presence of known or postulated
clusters in an unbiased fashion. For example in \cite{hyper}, the
\texttt{phenoDisco} algorithm was used to confirm the data-specific presence
of the nucleus and nucleus sub-compartments. In the code chunk below,
we demonstrate how to do this analysis, highlighting some of the
optional arguments and parameters available for phenotype extraction
and give some advice on how to interpret the output.

As the \texttt{phenoDisco} algorithm is semi-supervised it uses both
labelled (markers) and unlabelled data to explore the data structure
and find new sub-cellular data clusters. Thus the first step is to
define some input labelled data i.e. markers, that the algorithm will
use as input for the supervised learning aspect of the algorithm. As
described in \cite{hyper} we define a set of markers to use as input
for the analyses that cover well-known residents from three distinct
organelle structures; the mitochondria, plasma membrane and ER, and
from three well-known and abundant protein complexes; the proteasome
and two ribosomal subunits, 40S and 60S.  These input markers are
stored in the \texttt{phenoDisco.Input} \texttt{featureData} column of
\texttt{hl} and below set by \texttt{fcol = "phenoDisco.Input"}. We
can use the convenience accessor function \texttt{getMarkers} to print
out a table of the markers contained in this marker set. These initial
markers were manually curated using information from the UniProt
database, the Gene Ontology and the literature.

@@

The argument \texttt{times} indicates the number of times we run
unsupervied Gaussian Mixture Modelling before defining a new phenotype
cluster.  The recommended minimum and default value is 100. In the
above code chunk we increase the value to \texttt{times = 200} as we
have found for larger datasets (e.g. 5000+ proteins) a higher
\texttt{times} is requried for convergence. \texttt{GS} defines the
minimum number of proteins allowed per new data cluster and thus
heavily influences what type of new clusters are extracted. For
example, if a user is interested in the detection of small complexes
they may wish to use a small \texttt{GS = 10}, or \texttt{GS = 20} etc.
If they wish to detect larger, more abundant sub-cellular niches a
much higher \texttt{GS} would be preferable. Specifying a small
\texttt{GS} can be more time consuming than using a larger
\texttt{GS}, and there is a trade off between finding interesting
small complexes and those that may not be of interest as we find there
is a tendancy to find more noise when using a small \texttt{GS}
compared to using a higher one.
